---
title: Windowed rank functions
author: Irene Steves
date: '2020-04-17'
slug: window-rank-functions
categories: []
tags: []
---

In my [recent exploration of window functions](https://irene.rbind.io/post/window-functions/), I realized didn't really know the differences between rank functions. The [dplyr documentation](https://dplyr.tidyverse.org/reference/ranking.html) lists out six functions, of which I pretty much only use one (`row_number()`):

* `row_number()`
* `ntile()`
* `min_rank()`
* `dense_rank()`
* `percent_rank()`
* `cume_dist()`

Though the documentation description is relatively clear, it was still hard to grasp exactly how they differed. I found it easier to do the comparison visually. 

## Trivia score ranks

Given a toy dataset of trivia scores from two teams, let's see how the scores rank using the functions above.

```{r message = FALSE}
library(tidyverse)
theme_set(theme_minimal())

set.seed(20200417) #so I get the same teams each time
trivia_scores <- tibble(team = sample(c("cheetah", "ibex"), 10, replace = TRUE),
                        score = c(3, 6, 6, 18, 39, 40, 40, 40, 42, 99))

window_rank_example <- trivia_scores %>% 
    mutate(ntile3 = ntile(score, 3),
           min_rank = min_rank(score),
           dense_rank = dense_rank(score),
           percent_rank = percent_rank(score) %>% round(3),
           cume_dist = cume_dist(score)) 
```

```{r echo = FALSE, message = FALSE}
library(kableExtra)
library(knitr)

spec_color_set <- partial(spec_color, alpha = .7, begin = 1, end = .5,
                          option = "magma")

window_rank_example %>% 
    mutate_at(vars(ntile3, min_rank, dense_rank, percent_rank, cume_dist),
              ~cell_spec(.x, background = spec_color_set(.x))) %>% 
    kable(escape = F, align = "c") %>% 
    kable_styling(c("striped", "condensed"))
```

From the patterns above we see that:

* the distances between scores don't matter (39 to 40 is treated the same way as 42 to 99)
* all except for `ntile()` clump the three 40's together into one team
* `min_rank()` reaches 10 (the total number of scores) whereas `dense_rank()` reaches 7 (the total number of _distinct_ scores)
* `percent_rank()` is similar to `min_rank()` except that (a) it ranges from 0 to 1 and (b) each step is 0.111 instead of 1
* each step of `cume_dist()` is 1 / # scores (1/10 = 0.1 in this case) and is similar to `min_rank()` except that all ties get the _maximum_ score

To dig into it even more, here's another way of visualizing the ranks. Since ranks are only dependent on the _order_ of values and not the value itself (except with repeats), I've used row number for the x-axis and added it as a light gray background as well. 

We see that with the exception of `ntile`, all ranking functions group repeated numbers together (I colored the <span style = 'color:#81d5a6;font-weight:bold'>two 6's in seagreen</span> and the <span style = 'color:#ff8c69;font-weight:bold'>three 40's in salmon</span>). 

```{r echo = FALSE}
window_rank_example %>% 
    mutate(row = factor(row_number())) %>% 
    gather(type, value, -team, -score, -row) %>% 
    mutate(type = fct_inorder(type)) %>% 
    mutate(coloring = case_when(
        score == 6 ~ "six",
        score == 40 ~ "forty",
        TRUE ~ "other"
    )) %>% 
    filter(type %in% c("ntile3", "min_rank", "dense_rank")) %>% 
    ggplot(aes(row, value, fill = coloring)) +
    geom_col(aes(row, as.numeric(row)), fill = "grey90") +
    geom_col(size = 5) +
    facet_grid(~ type) +
    scale_y_continuous(n.breaks = 10, expand = c(0, 0)) +
    scale_fill_manual(values = c("six" = "#81d5a6", 
                                 "forty" = "#ff8c69",
                                 "other" = "grey60")) +
    guides(fill = FALSE) 
```

Since `percent_rank()` and `cume_dist()` reach a maximum of 1, I decided to plot them separately. The light gray background indicates equal increments of 0.1 across all 10 values.

```{r echo = FALSE}
window_rank_example %>% 
    mutate(row = factor(row_number())) %>% 
    gather(type, value, -team, -score, -row) %>% 
    mutate(type = fct_inorder(type)) %>% 
    mutate(coloring = case_when(
        score == 6 ~ "six",
        score == 40 ~ "forty",
        TRUE ~ "other"
    )) %>% 
    filter(type %in% c("percent_rank", "cume_dist")) %>% 
    ggplot(aes(row, value, fill = coloring)) +
    geom_col(aes(row, as.numeric(row) / 10), fill = "grey90") +
    geom_col(size = 5) +
    facet_grid(~ type) +
    scale_y_continuous(n.breaks = 10, expand = c(0, 0)) +
    scale_fill_manual(values = c("six" = "#81d5a6", 
                                 "forty" = "#ff8c69",
                                 "other" = "grey60")) +
    guides(fill = FALSE) 
```


## Ranking per group

Splitting by team requires just one extra line (and one more to make the resulting table look nicer):

```{r}
window_rank_example_teams <- trivia_scores %>% 
    group_by(team) %>% # only change needed!
    arrange(team, score) %>% # not required but easier to digest visually  
    mutate(ntile3 = ntile(score, 3),
           min_rank = min_rank(score),
           dense_rank = dense_rank(score),
           percent_rank = percent_rank(score) %>% round(3),
           cume_dist = cume_dist(score)) 
```

```{r echo = FALSE, message = FALSE}
window_rank_example_teams %>% 
    mutate_at(vars(ntile3, min_rank, dense_rank, percent_rank, cume_dist),
              ~cell_spec(.x, background = spec_color_set(.x))) %>% 
    kable(escape = F, align = "c") %>% 
    kable_styling(c("striped", "condensed"))
```

Now, instead of overall ranks, the scores for each team are ranked separately. 

## Reversing ranks

Perhaps you think it's weird that the lowest trivia scores get the highest rank (1). To reverse the ranking order, we can use either `desc()` or just add a `-`.

```{r message = FALSE}
window_rank_example_reversed <- trivia_scores %>% 
    mutate(ntile3 = ntile(-score, 3),
           min_rank = min_rank(-score),
           dense_rank = dense_rank(-score),
           percent_rank = percent_rank(-score) %>% round(3),
           cume_dist = cume_dist(-score)) 
```

```{r echo = FALSE, message = FALSE}
window_rank_example_reversed %>% 
    mutate_at(vars(ntile3, min_rank, dense_rank, percent_rank, cume_dist),
              ~cell_spec(.x, background = spec_color_set(.x))) %>% 
    kable(escape = F, align = "c") %>% 
    kable_styling(c("striped", "condensed"))
```

## Conclusion

And that's all there is to it! Hope the visual aids help to reinforce the differences in the options.
